\section{L4: Naive Bayes, Vectors, SGNS}
\subsection*{Multinomial Naive Bayes}
\[
\hat c_{MAP}=\arg\max_{c\in C}P(c\mid x).
\]
\[
\hat c_{MAP}=\arg\max_{c\in C}P(c)P(x\mid c).
\]
For bag-of-words counts $x_i$:
\[
P(x\mid c)=
\frac{n!}{\prod_{i=1}^{|V|}x_i!}
\prod_{i=1}^{|V|}\theta_{c,i}^{x_i}.
\]
\[
n=\sum_i x_i.
\]
Log-space score:
\[
\hat c=\arg\max_{c\in C}
\left[\log P(c)+\sum_{i=1}^{|V|}x_i\log\theta_{c,i}\right].
\]
\[
\theta_{c,i}^{MLE}=\frac{n_{i,c}}{n_c},
\qquad
\theta_{c,i}=\frac{n_{i,c}+\alpha}{n_c+\alpha|V|}.
\]

\subsection*{Binary NB vs Bernoulli NB (exam trap)}
Binary NB (lecture variant): clip counts at 1, then use multinomial-style training.\\
Bernoulli NB: models presence/absence explicitly per token dimension.\\
Both are "0/1-like", but not the same likelihood model.

\subsection*{Word Vectors and Similarity}
\[
a\cdot b=\sum_i a_i b_i.
\]
\[
\|a\|=\sqrt{\sum_i a_i^2},
\qquad
\cos(a,b)=\frac{a\cdot b}{\|a\|\|b\|}.
\]
Use cosine when length/frequency bias is undesirable.

\subsection*{Window Size and Neighbor Type (L4)}
Small window ($C=\pm2$): neighbors tend to be syntactic/taxonomic.\\
Large window ($C=\pm5$): neighbors become topical/semantic-field.

\subsection*{Analogy by Parallelogram}
\[
v(b^{*})\approx v(b)-v(a)+v(a^{*}).
\]
Examples: $v(king)-v(man)+v(woman)\approx v(queen)$,\\
$v(Paris)-v(France)+v(Italy)\approx v(Rome)$.\\
Works better for some relation types than others.

\subsection*{Embedding Bias Notes}
Embeddings reflect training statistics and social bias; downstream systems can inherit this bias.

\subsection*{SGNS Core (lecture form)}
\[
P(+\mid w,c)=\sigma(u_c\cdot v_w).
\]
\[
P(-\mid w,c)=\sigma(-u_c\cdot v_w).
\]
Training keeps word vectors $v_w$ and context vectors $u_c$.\\
Final representation often uses $v_w$ (or $v_w+u_w$).
\[
s^+=u_{c^+}\cdot v_w,\qquad s_i^-=u_{c_i^-}\cdot v_w.
\]
\[
L_{CE}= -\left[\log\sigma(s^+)+\sum_{i=1}^{k}\log\sigma(-s_i^-)\right].
\]
\[
\frac{d}{d\alpha}\log\sigma(\alpha)=\sigma(-\alpha).
\]
\[
\frac{d}{d\alpha}\log\sigma(-\alpha)=-\sigma(\alpha).
\]

\subsection*{MCQ Transfer Rules (no raw answer keys)}
If option mixes ranking with probability values in NB, prefer log-score form\\
$\log P(c)+\sum_i x_i\log\theta_{c,i}$ (same argmax, safer numerically).\\
For similarity questions: use cosine when vector norms/frequency should be removed;\\
dot product keeps magnitude bias.\\
For SGNS signs: positive pair uses $\log\sigma(+u_c\cdot v_w)$;\\
negative sample uses $\log\sigma(-u_{c^-}\cdot v_w)$.\\
Window-size questions: small window $\Rightarrow$ syntactic neighbors,\\
large window $\Rightarrow$ topical/semantic neighbors.\\
Analogy rule is approximate, not universal; expect reliability differences by relation type.

\subsection*{A1-style Formula Transfer}
Add-one bigram:
\[
P_{add1}(w_i\mid w_{i-1})=\frac{C(w_{i-1},w_i)+1}{C(w_{i-1})+|V|}.
\]
Interpolation template:
\[
P(w_i\mid h)=\lambda_1P_1+\lambda_2P_2,\qquad \lambda_1+\lambda_2=1.
\]
No-EOS caveat:
\[
\sum_{w_{1:\ell}}P(w_{1:\ell})=1\ \text{for fixed }\ell.
\]
Not globally normalized across all lengths.
NB derivation path: Bayes $\to$ drop class-independent terms $\to$ log-linear score.

\subsection*{L4 Pitfalls}
- Compute NB in log space to avoid underflow.\\
- Without smoothing, one unseen word can zero a class score.\\
- Dot product alone can over-favor frequent words; cosine is safer.
