\section{L1 Foundations: Tokens, Unicode, BPE, Regex}

\subsection*{Token Basics}
Type: unique vocabulary item; $|V|$.\\
Token (instance): occurrence in corpus; total $N$.\\
Orthographic $\neq$ grammatical word: ``I'm'' = 1 token, 2 words ($I+am$).\\
Heaps/Herdan law:
\[
|V| \approx K N^{\beta}, \qquad \beta \approx 0.5
\]
Vocabulary grows sublinearly; type-token ratio $|V|/N \downarrow$ as $N \uparrow$.\\
\textbf{Common traps:} token count depends on tokenizer; punctuation decisions matter.

\subsection*{Language \& Token Budget}
Whitespace tokenization not universal.\\
Chinese/Japanese/Thai require segmentation; multiple valid segmentations.\\
Token budget and perplexity not directly comparable across tokenizers/languages.\\
\textbf{Common traps:} tokenizer mismatch invalidates model/perplexity comparison.

\subsection*{Morphology and ``What is a Word''}
Morpheme: minimal meaning-bearing unit.\\
Root = core meaning; affix = grammatical/semantic modification.\\
Inflectional (e.g., \texttt{-ed}, \texttt{-s}) vs derivational (class/meaning change).\\
Clitic: syntactically word-like but phonologically attached (e.g., \texttt{'ve}).\\
\textbf{Common traps:} word boundary $\neq$ morpheme boundary.

\subsection*{Unicode \& UTF-8}
Unicode assigns code points (e.g., \texttt{a}=U+0061); code point $\neq$ glyph.\\
UTF-8: variable-length, ASCII-compatible.\\
ASCII (U+0000--U+007F) = 1 byte.\\
CJK typically 3 bytes; emoji often 4 bytes (range-dependent).\\
Python 3 strings = Unicode code points; \texttt{len()} counts code points (not bytes).\\
\textbf{Common traps:} mixing byte-level and code-point operations causes silent bugs.

\subsection*{BPE (Byte Pair Encoding)}
Two parts: (1) learner, (2) encoder.\\
Training (repeat $k$ merges):\\
\quad 1. Find most frequent adjacent pair $(A,B)$.\\
\quad 2. Add merged token $AB$ to $V$.\\
\quad 3. Replace occurrences of $A\ B$ with $AB$.\\
Encoding: start from chars/bytes; apply merges greedily in learned order.\\
Byte-level BPE: alphabet size 256 (all bytes) $\Rightarrow$ no OOV.\\
\textbf{Common traps:} test-time frequencies do NOT affect merge order.

\subsection*{Regex Essentials (Preprocessing)}
Character classes: \verb|[A-Z]|, \verb|[^0-9]|.\\
Quantifiers: \verb|*|, \verb|+|, \verb|?|, \verb|{m,n}|.\\
Anchors: \verb|^|, \verb|$|, boundary \verb|\b|, wildcard \verb|.|.\\
Grouping: capturing \verb|(...)|, non-capturing \verb|(?:...)|.\\
Lookahead: \verb|(?=...)| (pos), \verb|(?!...)| (neg).\\
\texttt{re.sub()} supports capture-group reordering.\\
\textbf{Common traps:} forgetting raw string \verb|r"... "| in Python.

\subsection*{Minimum Edit Distance (Levenshtein)}
Initialization:
\[
D(i,0)=i, \quad D(0,j)=j
\]
Recurrence:
\[
D(i,j)=\min
\begin{cases}
D(i-1,j)+1\\
D(i,j-1)+1\\
D(i-1,j-1)+
\begin{cases}
0 & x_i=y_j\\
2 & x_i\neq y_j
\end{cases}
\end{cases}
\]
Time $O(nm)$, space $O(nm)$ (can reduce to $O(\min(n,m))$).\\
Backtrace cost $O(n+m)$.\\
\textbf{Common traps:} substitution cost may be 2 (not 1).

\subsection*{Bigram \& N-gram Models}
Chain rule:
\[
P(w_1^N)=\prod_{i=1}^{N} P(w_i|w_1^{i-1})
\]
Bigram assumption:
\[
P(w_i|w_1^{i-1}) \approx P(w_i|w_{i-1})
\]
MLE:
\[
P(w_i|w_{i-1})=\frac{\text{count}(w_{i-1},w_i)}
{\text{count}(w_{i-1})}
\]
Add-one smoothing:
\[
\frac{\text{count}(w_{i-1},w_i)+1}
{\text{count}(w_{i-1})+|V|}
\]
Interpolation:
\[
P=\lambda_1 P_{\text{bigram}} + \lambda_2 P_{\text{unigram}}
\]
\textbf{Common traps:} include start/end symbols; MLE gives zero for unseen events.

\subsection*{Log-space \& Cross-Entropy}
Log trick: $\log(ab)=\log a + \log b$ (avoid underflow).\\
Cross-entropy (single example):
\[
-\log P(y|x)
\]
Maximize likelihood $\Leftrightarrow$ minimize cross-entropy.\\
\textbf{Common traps:} log does not change $\arg\max$ ranking.