\section{L2: Min Edit Distance, N-grams, Perplexity, Smoothing}
\subsection*{Min Edit Distance (Levenshtein / Minimum Edit Distance)}
Given strings $X=x_1\ldots x_M$, $Y=y_1\ldots y_N$:
\[
D(i,0)=i,\qquad D(0,j)=j.
\]
\[
\begin{array}{l}
D(i,j)=\min\{\\
\quad D(i-1,j)+1,\\
\quad D(i,j-1)+1,\\
\quad D(i-1,j-1)+cost\\
\}.
\end{array}
\]
$cost=0$ if $x_i=y_j$, else substitution cost (lecture examples may use $2$).\\
Weighted form:
\[
\begin{array}{l}
D(i,0)=D(i-1,0)+del[x_i],\\
D(0,j)=D(0,j-1)+ins[y_j],\\
D(i,j)=\min\{quad D(i-1,j)+del[x_i],\\
\quad D(i,j-1)+ins[y_j],\\
\quad D(i-1,j-1)+sub[x_i,y_j]}.
\end{array}
\]
Backtrace: store operation pointer (DEL/INS/SUB-MATCH) per cell.\\
Complexity: time $O(MN)$; memory $O(MN)$ or $O(\min\{M,N\})$ with rolling rows.

\subsection*{DP + Backtrace Mini-Template}
Dynamic programming fill order: row 0/col 0 first, then left-to-right top-to-bottom.\\
Cell transition compares \texttt{DEL/INS/SUB(or MATCH)} candidates; keep minimum cost and pointer.\\
Alignment reconstruction: start at $(M,N)$ and follow pointers to $(0,0)$; reverse operations.\\
Weighted edit distance uses confusion-informed costs (keyboard typo/ASR confusions).

\subsection*{N-gram Language Modeling}
Chain rule:
\[
P(w_1^n)=\prod_{i=1}^{n}P(w_i\mid w_1^{i-1}).
\]
Markov assumption (order $N-1$):
\[
P(w_i\mid w_1^{i-1})\approx P(w_i\mid w_{i-(N-1)}^{i-1}).
\]
N-gram approximation:
\[
P(w_i\mid w_1^{i-1})\approx P(w_i\mid w_{i-(N-1)}^{i-1}).
\]
MLE bigram/unigram:
\[
P_{ML}(w_i\mid w_{i-1})=\frac{C(w_{i-1},w_i)}{C(w_{i-1})},
\]
\[
P_{ML}(w_i)=\frac{C(w_i)}{\sum_{w\in V}C(w)}.
\]
With sentence markers:
\[
P(w_1\ldots w_n)=\prod_{i=1}^{n}P(w_i\mid w_{i-1}).
\]
\[
w_0=\langle s\rangle.
\]
If EOS is modeled, terminate with \texttt{</s>} and include EOS counts in training.

\subsection*{Train/Dev/Test Protocol (L2)}
Train set: fit parameters.\\
Dev set: tune hyperparameters/model choices.\\
Test set: final one-shot report of generalization.\\
Lecture warning: training (or repeated tuning) on test set gives\\
artificially high test probability; bad science.\\
Workflow: train $\rightarrow$ tune on dev $\rightarrow$ freeze model $\rightarrow$ run test once.\\
Generalization check: lower train PP but higher dev/test PP indicates overfitting.

\subsection*{Perplexity (textbook canonical)}
For test sequence $W=w_1\ldots w_N$:
\[
PP(W)=P(w_1\ldots w_N)^{-1/N}.
\]
\[
PP(W)=\exp\!\left(-\frac{1}{N}\sum_{i=1}^{N}\log p_i\right).
\]
\[
p_i=P(w_i\mid w_1^{i-1}).
\]
\[
PP(W)=2^{H(W)}.
\]
L2 intuition slides: perplexity is inverse normalized test probability;\\
minimizing PP is equivalent to maximizing test probability.\\
Interpretation: weighted average branching factor (effective number of likely next words).\\
Comparison rule: only compare PP under same tokenizer/preprocessing/test split.

\subsection*{Shannon Visualization (1948)}
Shannon game intuition: a good LM predicts the actually occurring next word with higher probability.\\
Bigram sampling procedure (slide style):\\
1. Sample first word $w$ from $p(w\mid\langle s\rangle)$.\\
2. Repeatedly sample next token $x\sim p(x\mid w)$.\\
3. Set $w\leftarrow x$ and continue until \texttt{</s>}.\\
4. Concatenate sampled words to form generated sentence.

Lower PP is better only under same tokenizer, preprocessing, and test split.

\subsection*{Smoothing / Backoff / Interpolation}
Laplace (add-1):
\[
P_{add1}(w_i\mid w_{i-1})=
\frac{C(w_{i-1},w_i)+1}{C(w_{i-1})+|V|}.
\]
Add-$k$ smoothing (\texttt{add-k}, $0<k<1$):
\[
P_{addk}(w_i\mid w_{i-1})=
\frac{C(w_{i-1},w_i)+k}{C(w_{i-1})+k|V|}.
\]
Trigram add-one (exercise form):
\[
P_{add1}(w_3\mid w_1,w_2)=
\frac{C(w_1,w_2,w_3)+1}{C(w_1,w_2)+|V|}.
\]
Linear interpolation (trigram):
\[
\begin{array}{l}
P(w_i\mid w_{i-2},w_{i-1})=\\
\lambda_3P_{ML}(w_i\mid w_{i-2},w_{i-1})+\\
\lambda_2P_{ML}(w_i\mid w_{i-1})+\lambda_1P_{ML}(w_i),
\end{array}
\]
\[
\lambda_1+\lambda_2+\lambda_3=1.
\]
Backoff principle: when higher-order count is zero, use lower-order model (usually with discounting).

\subsection*{Exercise-Style Quick Checks (Ch2/3)}
3.2/3.3 style: unsmoothed MLE often gives sharper in-corpus probabilities;\\
smoothing redistributes mass to unseen events.\\
3.4/3.7 style: \texttt{I am Sam} corpus yields $P(Sam\mid am)=\frac{3}{14}$ (add-1)\\
and $\frac{31}{75}$ (interpolation).\\
3.10/3.11 style: sampling and perplexity computation should use exactly the same LM definition/tokenization.

\subsection*{A1 Repro (L2 core)}
Q2(a) add-one bigram:
\[
P(Sam\mid am)=\frac{C(am,Sam)+1}{C(am)+|V|}=\frac{3}{14}.
\]
Q2(b) interpolation in assignment:
\[
P(Sam\mid am)=\frac{31}{75}.
\]
Q3 no EOS: for fixed length $\ell$,
\[
\sum_{w_{1:\ell}}P(w_{1:\ell})=1,
\]
but across all lengths it is not normalizable.

\subsection*{L2 Pitfalls}
- [YiCuoDian] Unsmoothed model assigns zero probability to unseen n-grams.\\
- Add-1 is often too aggressive for large vocabularies; add-$k$/interpolation is often stabler.\\
- Tune interpolation weights on held-out dev data, not test set.\\
- Always include/exclude \texttt{</s>} consistently across training and evaluation.
